---
title: "Introduction to R"
author: "Kristian Brock"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(dplyr)

knitr::opts_chunk$set(echo = TRUE)

abc <- c(1, 4.5, 99, 783.276, 10 / 3)

more_trials <- data.frame(
  name = c('PePS2', 'BTOG2', 'Viola', 'RomiCar', 'Matrix', 'Trapeze'),
  disease = c('lung', 'lung', 'haem', 'haem', 'lung', 'prostate'),
  n = c(62, 1363, 29, 20, 200, 707),
  arms = c(1, 3, 1, 1, 20, 2),
  phase = ordered(c('II', 'III', 'I', 'I', 'II', 'III'), 
                  levels = c('I', 'II', 'III'))
)
```


## Welcome

This tutorial introduces you to R and RStudio.
It intends to get you up and running.


### Learning Objectives
Upon completion of this session, you will be able to: 

* Use RStudio to run commands in base R and tidyverse R;
* Work with RStudio projects;
* Install and use the crctu R-package



## Base R

At its simplest, R can be used as a calculator.
You run commands and R tells you the answer.
For example, if you type:

```{r}
365 / 7
```
you learn that there are slightly more than 52 weeks in a year.


### Vector
One of the core objects in R is a `vector`.
It collects instances of the same type together.
You can create a vector using the `c()` function.
For instance, create a vector of numbers using code like:

```{r}
abc <- c(1, 4.5, 99, 783.276, 10 / 3)
```

The `<-` operator assigns data to a name of your choosing.
I chose `abc`. 
You might instead have chosen `x` or `data`, etc.

Type the assignment operator using the `<` character followed by `-`.
Assignment in R actually works in both directions.
The chunk above could also have been written:

```{r}
c(1, 4.5, 99, 783.276, 10 / 3) -> abc
```

Back to vectors.
There are many functions in R that operate on vectors.
For example, to calculate the sample mean and standard deviation of these numbers, you would run:

```{r}
mean(abc)
```

and 

```{r}
sd(abc)
```

Your turn. 
Calculate the median of `abc`.
Type in the command then hit CTRL + Return (on Windows) or Command + Return (on Mac) to execute the command.


```{r median, exercise = TRUE}

```

Notice the intellisence trying to guess which function you were aiming for?
Hitting Return will autocomplete the highlighted suggestion.
RStudio and R are very helpful like that.
If it irritates you, hit Escape to kill it.

Did your median calculation produce the answer you were expecting?

Commands like `mean` and `median` take a vector and reduce it to a single summary statistic.
In contrast, there are many commands in R that are _vectorised_, i.e. they perform an operation on each element of a vector.
This is best understood with an example.
What do you imagine this does:
```{r vectorise-1, exercise = TRUE}
abc + 2
```
Run it and find out.
Is that what you guessed?

And what do you think is going on here:
```{r vectorise-2}
abc + abc^2
```

There are many functions that operate on vectors in R and return either a single statistic or another `vector`, including `length`, `sum`, `abs`, `cos`, `exp`, `log`, `log10`, .... etc.
Have an explore here:

```{r explore-functions, exercise = TRUE}

```


You can access the elements of a `vector` using square brackets and the integer index.
The first element is:
```{r}
abc[1]
```

R is an _object oriented_ language so everything has a type.
You can check the type of an object using the `class` command.
For example:

```{r}
class(abc)
```

```{r}
class(c('The', 'quick', 'brown', 'fox'))
```

Remember, the objects in a `vector` are all of the same type.
The same is not true of lists.

### List
Another useful data-type is the `list`.
Whereas vectors contained multiple instances of the same type, lists can contain values of different types.
For example:

```{r}
l <- list(Name = "Kristian", Weight = 75, JoinedCRCTU = as.Date('2014-02-10'))
l
```

This list contains textual, numerical and date data.
You access elements of a list using double square brackets and integer indices:
```{r}
l[[1]]
```

or textual names:
```{r}
l[['JoinedCRCTU']]
```

Lists can contain essentially anything as an element, including vectors and even other lists.
There are functions that operate on lists that we will encounter later.


### Data-frame
The final data-type that you must know about is the data-frame, written in R as `data.frame`.
These can be throught of as spreadsheets of data arranged in columns and rows.
Each column has a type and a name.
Rows can have names (but they need not).

There are many ways of creating data-frames but here is perhaps the simplest:

```{r}
trials <- data.frame(
  name = c('PePS2', 'Matrix', 'Viola'),
  phase = ordered(c('II', 'II', 'I'), levels = c('I', 'II', 'III'))
)

trials
```

This example introduces the `ordered` data-type.
This is a special type of factor variable where the levels are ordered.

Extend the example above to add a variable called `n` containing the following trial sample sizes:

* PePS2 has 62 patients;
* Matrix has 200 patients;
* Viola has 24 patients.

```{r extend-dataframe, exercise = TRUE, exercise.eval = FALSE}
data.frame(
  
) -> trials

trials
```

```{r extend-dataframe-hint}
data.frame(
  trials, 
  n = c()
) -> trials

trials
```

We will learn about a rich grammar for working with data-frames in the next Section.


## Tidyverse

The `tidyverse` is a set of packages developed by RStudio that make working with data in R simple and intuitive, at least in the opinion of this author.
The jewel in the `tidyverse` crown is `dplyr`, a package for working with data-frames.
It is not part of base R and so must be loaded explicitly using the `library` command:

```{r, message=FALSE, warning=FALSE}
library(dplyr)
```

We will take our `trials` example from the previous Section and expand it a little:
```{r}
more_trials <- data.frame(
  name = c('PePS2', 'BTOG2', 'Viola', 'RomiCar', 'Matrix', 'Trapeze'),
  disease = c('lung', 'lung', 'haem', 'haem', 'lung', 'prostate'),
  n = c(62, 1363, 29, 20, 200, 707),
  arms = c(1, 3, 1, 1, 20, 2),
  phase = ordered(c('II', 'III', 'I', 'I', 'II', 'III'), 
                  levels = c('I', 'II', 'III'))
)

more_trials
```

One of the hallmarks of tidyverse R is the use of `%>%`, the pipe operator.
It means _take what is on the left and use it as the first parameter in the function on the right_.
Put like that, it doesn't sound particularly useful.
Some examples might help.
Let's retain only trials with `n` greater than 100:

```{r}
more_trials %>% 
  filter(n > 100)
```

Your turn.
Write a command to show only trials in `haem`:
```{r just_haem, exercise = TRUE}

```

```{r just_haem-hint}
# Test equality of a column containing text using
filter(col_name == 'sought_value')
```


Another way of writing the command that filtered on sample size is:
```{r}
filter(more_trials, n > 100)
```

However, using the `%>%` operator is powerful because it allows you daisy-chain commands together to create complex analyses.
For example, we can calculate the number of patients per arm in trials with a sample size of at least 100:

```{r}
more_trials %>% 
  filter(n > 100) %>% 
  mutate(pats_per_arm = n / arms)
```

A very useful combination of verbs is `group_by` and `summarise`.
For example, let's say we are interested in analysing sample size by trial phase.
We could proceed:

```{r}
more_trials %>% 
  group_by(phase) %>% 
  summarise(
    min_n = min(n),
    avg_n = mean(n),
    max_n = max(n)
  )
```

That looks about right to me.
Now you do some examples, pooling verbs from the previous examples.

What is the average sample size by disease area?
```{r n-by-disease, exercise = TRUE}

```

What is the average number of arms by disease area, excluding trials in phase I?

```{r arms-by-disease, exercise = TRUE}

```

```{r arms-by-disease-hint}
# remove phase I trials in this example using one of:
filter(phase != 'I')
filter(phase == 'II' | phase == 'III')
filter(phase %in% c('II', 'III'))
```


## RStudio
The R application is a little austere, in that it provides few luxuries.
Most people interact with R through another program and the most common is probably RStudio.

RStudio makes it easier to write R code, run code, write documentation and research articles, interact with source control. 
We recommend it without hesitation.

### Desirable changes
There are, however, some things we request that RStudio users do in CRCTU to promote reproducible research and responsible coding for others.
Please take the following steps on any RStudio instance you use for CRCTU work.

1. Select "Tools -> Global Options…". Under "General", please:
  a) set "Save workspace to .Rdata on exit" to "Never";
  b) uncheck "Restore .Rdata into workspace at startup".

Why would we do this?

Please also:

2. Select "Tools -> Global Options…". Under "General":
  a) 	Under "Code -> Display", check "Show margin" and set "Margin column" to 80. 
  
This creates a vertical line at 80 characters. 
When writing code, please do not exceed this margin. 
Nobody should have to scroll across to read what you have written. 
  
The code you and I write belongs to CRCTU and it will need to be executed by others. 
This is the first step to good coding style that will promote writing software that others can run and maintain.
There are many other elements of coding style we should follow as well.
Any volunteers to research and develop this?

### Make this place your own

There are other steps you should explore to make your RStudio session your own.

Under "Tools -> Global Options", experiment with different pane layouts. 
I personally like to have code editor in top-left pane and console output on top-right.
That makes sense to me but you might prefer something different.

You never really live somewhere until you have decorated, so change the RStudio theme.
Under "Tools -> Global Options", go to the Appearance tab and try out some different editor themes.
I quite like some of the dark themes like Cobalt and Vibrant Ink.

Extra special brownie points are available to anyone with the tekkers to create their own theme.
I hope you are tempted down this road.
I am but I have never found the time.


## Packages
R packages are a way of grouping together R code, help files, long-form documentation called _vignettes_, tutorials (just like this one), and document templates.

Packages are an essential part of using R because base R is actually fairly limited. 
Many of the useful capabilities of R are provided by packages.
For instance, we saw in the previous examples that the commands `filter`, `mutate`, `group_by`, `summarise` and many others are provided by the `dplyr` package.
This is fairly typical of open-source languages.
Both Julia and Python take a similar approach.

Typically, packages are installed from CRAN (Comprehensive R Archive Network), the official repository for R packages.
Packages can be installed from elsewhere too, like GitHub, or a local zip file.
Ultimately, someone needs to write the package and make it available.

### Libraries of packages and CRCTU
R installs packages to a _library_.
The default library location is a sub-directory of the user's home path.
Historically, this has created problems in CRCTU with packages that use C++ compilation like `rstan`.
Other problems have arisen regarding use of our (networked) home drives.
If you ever run into problems with the library location, you can change the location manually by following these steps:

* Create the folder C:\\R\\Library.
* Click Start --> Control Panel --> User Accounts --> Change my environmental variables.
* The Environmental Variables window pops up. ...
* Change or add the R_LIBS_USER variable so that it points to the directory that you created. 

You will not need to do this on the RStudio Cloud instance we use for `bootcamp` or CRCTU's RStudio Server instance.

### Install a package from CRAN
The primary source for R packages is CRAN.
To install a package from CRAN, you run:
```{r, eval=FALSE}
install.packages('package_name')
```

This fetches the latest version of `package_name` and installs it in the library of packages on your system.
Crucially, it also tries to satisfy all package dependencies.
Packages depend on other packages to perform tasks for them.
This allows packages to specialise in precise tasks and avoids the need to reinvent the wheel many times.
It does, however, create a network of package dependencies that can sometimes be overwhelming to behold.

**Exercise**: Look up the CRAN page for the `lme4` package. What packages does it rely on?


If a package you are trying to install requires a newer version of a package dependency than that present on your system, it may try to upgrade without warning you.
This might have adverse consequences for the reproducibility of previous analyses you have run.
Thus, a little bit of care is needed when installing packages.
Be aware of dependencies and try to anticipate whether upgrades will present a problem.
Nevertheless, software moves relentlessly forwards and a pragmatic embracing of upgrades will often by inevitable.
For example, how many apps have updated on your iOS or Android phone in the last week? 

**Exercise**: What does the R package `MASS` do? When was the last version released? What changes did the last release bring? Install the package.
```{r install-MASS, exercise = TRUE}

```

You can test that `MASS` is now installed by running: 
```{r library-MASS, exercise = TRUE}
library(MASS)
```

If the package is present, this command should execute with no error.

### Install a package from GitHub
In many instances, packages are available on GitHub as well as CRAN.
Typically, bleedy-edge releases (i.e. those not yet added to CRAN) can be obtained by installing directly from GitHub.
However, you should bear in mind that you might be installing untested or even non-working code. 
You have to use your common sense.

You can install from GitHub using the `devtools` package.
Execute the following:
```{r install-welcome, exercise = TRUE}
devtools::install_github('brockk/welcome')
```

What function(s) does the `welcome` package provide?
Type `welcome::` in the dialogue below and autocomplete should suggest functions.
Hit Return to accept the suggestion.
What does the function do?
```{r use-welcome, exercise = TRUE}

```

More complicated examples are available (obviously).
But you have installed and used a package from GitHub.
From time to time, that will be useful.


### Install from local zip file
You can also install packages from a zip file.
For instance, when packages fall foul of CRAN's evolving standards and rules, they are delisted from the `install.packages` route.
However, to promote reproducibility, the historic releases are available for manual download as a zip file from the CRAN page.
After download, these can be installed manually.

There are other cases where you might prefer to install a package from a local zip file.
For instance, if the code in a package is sensitive and you do not want to share it with the outside world.
In fact, this is precisely the situation with the `crctu` package.
It contains functions for performing common tasks at CRCTU like taking, saving and loading database snapshots.
It also contains template RMarkdown reports for various tasks - we will encounter those later in bootcamp.

If you are within CRCTU's firewall, you can install the `crctu` package from zip file.

Identify the most recent version at:

> T:\\Code\\R\\Packages\\crctu_X.Y.Z.tar.gz

Install it by selecting:

1. Tools -> Install packages...
2. Under Install from: select Package Archive File (.zip; .tar.gz)
3. A dialogue box should appear. Navigate to the file you identified above and select Open
4. Check the value of Install to Library: Is this right?
5. Hit Install
6. Test you can load it by running `library(crctu)`.

We will learn how to use `crctu` in a later class.


## Quiz

Let us recap some of what we have learned with a quiz.

```{r quiz, echo = FALSE}
quiz(
  
  question("Vectors allow values of different types to be associated?",
    answer("FALSE", correct = TRUE),
    answer("TRUE", correct = FALSE)
  ),

  question("Lists allow values of different types to be associated?",
    answer("FALSE", correct = FALSE),
    answer("TRUE", correct = TRUE)
  ),
  
  question("Data-frames can contain columns of different types",
    answer("FALSE", correct = FALSE),
    answer("TRUE", correct = TRUE)
  ),
  
  question("How would you remove unwanted rows in a data-frame using dplyr?",
    answer("summarise", correct = FALSE),
    answer("select", correct = FALSE),
    answer("filter", correct = TRUE),
    answer("mutate", correct = FALSE)
  ),
  
  question("How would you calculate statistics for each level of a group using dplyr?",
    answer("summarise", correct = TRUE),
    answer("select", correct = FALSE),
    answer("filter", correct = FALSE),
    answer("mutate", correct = FALSE)
  ),
  
  question("Which commands add new columns using dplyr?",
    answer("summarise", correct = TRUE),
    answer("select", correct = FALSE),
    answer("filter", correct = FALSE),
    answer("mutate", correct = TRUE)
  ),
  
  question("Code in packages on GitHub is always reliable.",
    answer("FALSE", correct = TRUE),
    answer("TRUE", correct = FALSE)
  ),
  
  question("Code in packages on CRAN is always reliable.",
    answer("FALSE", correct = TRUE),
    answer("TRUE", correct = FALSE)
  )
)
```


## Next steps

The material in this session forms the foundation of much of the future training sessions in R, in particular working with `ggplot2` and automating document creation using `rmarkdown`.

If you enjoyed using R and RStudio, please keep at it. 
There is nothing like using a tool to solve real problems to really learn how it works.

There are many many R resources online.

* A key reference is the book _R for Data Science_ by RStudio's Hadley Wickham.
Hadley wrote much of the `tidyverse` packages.
The book is available free online in its entirety at https://r4ds.had.co.nz/

The `learnr` tutorials are great:

* https://jjallaire.shinyapps.io/learnr-tutorial-01-data-basics/
* https://jjallaire.shinyapps.io/learnr-tutorial-03a-data-manip-filter/
* https://jjallaire.shinyapps.io/learnr-tutorial-03b-data-manip-mutate/
* https://jjallaire.shinyapps.io/learnr-tutorial-03c-data-manip-summarise/

There is an R package for learning R (obviously).
It is called `swirl`.
Launch an interactive session by running the following in RStudio:

```{r eval=FALSE}
install.packages('swirl')
library(swirl)
swirl()
```

There are further resources at www.codeacademy.com and www.datacamp.com


Please tell us what you find helpful.

Kristian Brock, CRCTU, 2019.
