---
title: "Using ggplot2"
author: "Kristian Brock"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE)

library(MASS)
library(dplyr)
library(ggplot2)

set.seed(123)
diamonds1k <- diamonds %>% sample_n(1000)

n_pat <- 50
times <- c(0, 3, 6, 9)
n_time <- length(times)

mu1 <- c(5, 4.5, 4, 3.8)
mu2 <- c(5, 4.9, 4.8, 4.8)
Sigma1 <- matrix(c(1, 0.9, 0.8, 0.7,
                   0.9, 1, 0.9, 0.8,
                   0.8, 0.9, 1, 0.9,
                   0.7, 0.8, 0.9, 1), ncol = length(mu1), byrow = TRUE)
set.seed(123)
X1 <- MASS::mvrnorm(n = n_pat, mu = mu1, Sigma = Sigma1)
X2 <- MASS::mvrnorm(n = n_pat, mu = mu2, Sigma = Sigma1)
df1 <- data.frame(
  TNO = rep(1:n_pat, n_time),
  TumourSize = as.numeric(X1),
  Time = rep(times, each = n_pat),
  Arm = 1
)
df2 <- data.frame(
  TNO = rep(n_pat + 1:n_pat, n_time),
  TumourSize = as.numeric(X2),
  Time = rep(times, each = n_pat),
  Arm = 2
)
tumours <- bind_rows(df1, df2)
```


## Welcome

This tutorial introduces you to the R package `ggplot2`.
It is an implementation of the _The Grammar of Graphics_ (Willis & Rope, Springer, 2005), a conceptual description of how to marry commands to data to produce practically every kind of visualisation seen in scientific publications.

### Learning Objectives
Upon completion of this session, you will:

* Be able to produce plots using ggplot2 in R; 
* Understand geometries and aesthetics in the grammar of graphics;
* Be able to visualise repeated measures data from RCTs.


## Geometries & aesthetics

The first step to making sense of your data is to visualise it.
According to the help-files of the `MASS` package, the `mcycle` data set contains:

> a series of measurements of head acceleration in a simulated motorcycle accident, used to test crash helmets.

Let's take a look at the data.

```{r}
mcycle %>% head()
```

We have two columns.
`times` contains the time in milliseconds after impact, and `accel` contains acceleraton measured in g.

Producing a scatter plot of these two variables in base R is trivially simple.
We simple call `plot` specifying the $x$ and $y$ arguments:

```{r}
plot(x = mcycle$times, y = mcycle$accel)
```

The plot may be simple but it is not particularly pretty.
One of the attractions of `ggplot2` is that the plots look journal-ready right from the outset.
To reproduce the above plot in `ggplot2`, we first load the package

```{r}
library(ggplot2)
```

and then run:

```{r}
mcycle %>% 
  ggplot(aes(x = times, y = accel)) + 
  geom_point()
```

The new plot is certainly prettier, but the commands to produce it are much more complex.
`ggplot2` certainly has a learning curve and may seem needlessly arcane initially.
But its power comes with its breadth and its consistency.
Once you learn the principles, you will quickly learn how to create all manner of plots.

The first thing to stress is that the command starts with the data and then proceeds with the pipe operator, `%>%`.
Recall that the pipe operator takes the object on the left and feeds it in as the first argument to the function on the right.
So the call above is equivalent to

```{r, eval = FALSE}
ggplot(mcycle, aes(x = times, y = accel)) + 
geom_point()
```

The second thing to stress is that plots in `ggplot2` are created with a call to the function `ggplot`.
Within this function is a call to another function called function `aes`.
This function lets you specify the _aesthetics_ of the plot.
Aesthetics in the grammer of graphics are how you associate variables in your data to properties of the plot.
The call to `aes` above is saying "The $x$-coordinates for this plot are stored in the `times` column of my data set, and the $y$-coordinates are stored in `accel`".
There are many aesthetics that can be specified, like `colour`, `group`, `fill`, `shape`, `linetype`. 
We will encounter these below.

The third thing to stress is that calls to `ggplot` are followed by calls to any number of _geometries_.
Each geometry adds a layer to the plot representing the data in some way.
Our example above uses `geom_point` which is the geometry for dropping points on a plot, as in a scatter plot.

You can add any number of geometries to a plot.
They are overlaid, like layers.
To add another to our example above, you simply append it using the `+` operator.
For example, if we wanted to join the dots in this plot, we can add a line using `geom_line`: 

```{r}
mcycle %>% 
  ggplot(aes(x = times, y = accel)) + 
  geom_point() + 
  geom_line()
```

Hopefully, the problem with this geometry in this context is obvious straight away.
The `accel` measurements are subject to error so the line geometry oscillates up and down, reflecting the noise about some underlying, imprecisely-observed process value.
Perhaps a more useful geometry would be one that smooths out this random variation like a [loess smoother](https://en.wikipedia.org/wiki/Local_regression).
Luckily, `ggplot2` provides such a function straight out of the box:

```{r}
mcycle %>% 
  ggplot(aes(x = times, y = accel)) + 
  geom_point() + 
  geom_smooth(method = 'loess', span = 0.4)
```

That is better.
We could easily read off the expected value at any time. 
We even get 95% confidence intervals for the mean thrown in for free.
In the example above I specified the `method` and `span`.
If you omit these details, `geom_smooth` will usually make excellent guesses at appropriate parameter values for you.
In this instance,however, manual tweaking was desirable.

So far, you have learned:

* plots are produced using calls to the `ggplot` function;
* aesthetics are conveyed using the `aes` function;
* geometries are overlaid to produce rich visualisations.

We will learn more aesthetics and geomtries in the following sections.

## Groups
We have learned how to put some points on a plot and join them up or fit a smooth line.
So far, so good.
Frequently, however, we will encounter data that is grouped in some way.
For instance, in a clinical trial we may have patients allocated to different treatments.
In this section we will learn how to reflect group-level information in `ggplot2` using:

* colour in plots;
* the `group_by` command;
* the `facet_wrap` and `facet_grid` commands.

We will use the `diamonds1k` dataset, a randomly-sampled 1000 rows from the `diamonds` dataset that ships with `ggplot2`.
The original dataset has 50,000 records.
We use a subset here because it is much faster.

```{r}
diamonds1k %>% head()
```

We see that the dataset includes information on the `carat` size and `price` of many diamonds, and that these are stored as numerical variables.
Furthermore, there are ordinal variables reflecting `cut`, `color`, and `clarity`.
Many of the online `ggplot2` examples use the `diamonds` dataset so building some familiarity with it will be beneficial.

### Colour
Let's examine the relationship between `carat` and `price`.
Using your skills from the previous section, produce a simple scatter plot of `carat` on the $x$-axis, and `price` on the $y$-axis.

```{r carat-vs-price, exercise = TRUE}
diamonds1k %>% 
  ggplot()
```

Describe the relationship.
If you want a bigger carat, what are you going to need?

We want to learn how to reflect group or cohort information in plots.
In many cases, merely using colour to distinguish one group from another will be sufficient. 

Let's investigate how `cut` affects the `carat`-`price` relationship using colour to distinguish the groups.
To produce a scatter-plot with a different colour for each level of `cut`, you need to tell `ggplot2` that the choice for colour will be determined by the `cut` variable.
Furthermore, because one of the aesthetics in your plot will be determined by one of your variables, the association between `colour` and `cut` is defined within a call to `aes`.
Have a go at producing that plot (There is a hint if you get stuck).

```{r carat-vs-price-by-cut, exercise = TRUE}


```

```{r carat-vs-price-by-cut-hint}
aes(x = ..., y = ..., col = cut)
```

Pretty.
Can you discern the effect that `cut` has on the relarionship between carat and price?
Me neither.
Giving each group its own subplot might make things clearer.
We cover that next.

### Facets
If your plot is a bit too noisy, sometimes a very useful trick is to make several versions of the same plot, with each showing the data for one particular group.
This is achieved in `ggplot2` using the `facet_wrap` and `facet_grid` commands.
These are applied to a plot using the `+` operator, exactly as new geometries are.
The subplots are displayed in a matrix shape.

`facet_wrap` takes as its first argument a command like `~ group_level`.
This tells it to produce a copy of the plot for each level of `group_level`.
For example, to facet our plot above by `cut`, we run:

```{r}
diamonds1k %>% 
  ggplot(aes(x = carat, y = price)) + 
  geom_point(aes(col = cut)) + 
  facet_wrap(~ cut)
```

Notice that the subplots are on identical axes?
This makes it possible to read horizontally and vertically to compare groups.

`facet_wrap` also takes optional parameters `ncol` or `nrow`, specifying the number of columns or rows to use in the grid of plots.
You only need to use one because the other is implicit.
If both are omitted, `ggplot2` decides for you.
Edit the example above to produce a plot with five columns:

```{r carat-vs-price-pivot-cut, exercise = TRUE}

  
```

From this plot, can you detect meaningful differences in expected `price` for different qualities of `cut`?
If not, maybe a smooth line fit to each group will make things clearer.
Recalling the previous page, we can add a smoother using `+ geom_smooth()`.
Do that now:

```{r carat-vs-price-pivot-cut-smooth, exercise = TRUE}


```

The smooths reflect only the data in each subplot.
This is because the `facet_wrap` command splits the dataset and that split perpetuates through to the call to `geom_smooth`.
Thus, grouping has been achieved without it really being that explicit.

What do you make of the relationship between `cut` and `price`?
The effect is detectable in this plot, but it looks surprisingly weak to me.

What do you expect a 1 carat diamond to cost for each type of cut?
To make it easier to look up from the $x$-axis, you can add a vertical line to each plot.
This is just another geometry, `geom_vline` to be precise.
And it is added in the usual way:

```{r carat-vs-price-pivot-cut-smooth-vline}
diamonds1k %>% 
  ggplot(aes(x = carat, y = price)) + 
  geom_point(aes(col = cut)) + 
  facet_wrap(~ cut, ncol = 5) + 
  geom_smooth() + 
  geom_vline(xintercept = 1, col = 'red', linetype = 'dashed')
```

Notice from the call to `geom_vline` that variables can be allocated values outside of `aes` when the values they take do not depend on the data.
For example, in this example we wanted the vertical line at $x = 1$, irrespective the data.
In another example where the location of the vertical line depended on the dataset, the  `xintercept` variable would instead have been allocated inside `aes` one of the columns in `diamonds1k`.

So let us revisit the question of interest: what do you expect a 1 carat diamond to cost for each type of cut?

Perhaps the effect of one of the other qualitative variables will be more pronounced.
Take the example immediately above and change it to reflect the effect of `clarity`:
```{r carat-vs-price-pivot-clarity-smooth, exercise = TRUE}




```


### Grouping

The third method of grouping data is to explicitly specify the `group` aesthetic.
This is required, for instance, when you want to show lines on the same plot.
A common example of this in medical research is repeated outcome measures, like quality of life scores.
We investigate such a scenario now with a simulated dataset from a two-arm trial where tumour size was measured at baseline and several times post-baseline.

The dataset is quite large but the first few rows look like this:
```{r}
tumours %>% head
```

The standard way of viewing the tumour sizes through time is:

```{r}
tumours %>% 
  ggplot(aes(x = Time, y = TumourSize)) + 
  geom_point()
```

but this plot shows no grouping information: we do not know which points join up. i.e. which points are associated with which patient.
Edit this example by appending `geom_line()` to get line-plots (do not worry about trying to convey the grouping variable for now):

```{r tumour-lines, exercise = TRUE, exercise.lines = 5}
tumours %>% 
  ggplot(aes(x = Time, y = TumourSize)) + 
  geom_point()
```

Does this look strange weird?
What has gone wrong?

The plot has joined the points as they arrived in the data-frame.
The result is that unrelated points have been joined together to create a non-sensical plot.
We need to set the `group` aesthetic within the call to `geom_line` to be `TNO` to reflect that only points within the same patient if (called `TNO`) should be joined.
We do that like this:

```{r tumour-lines-grouped, exercise = FALSE, exercise.lines = 5}
tumours %>% 
  ggplot(aes(x = Time, y = TumourSize)) + 
  geom_point() + 
  geom_line(aes(group = TNO))
```

Notice that there are now two calls to `aes`.
Aesthetics can be assigned inside the `ggplot` function and inherited by all geometries.
Or they can be overridden in a specific geometry by nesting a call to `aes` inside that `geom_XYZ` function.
Colour the points and lines by adding `col = TumourSize` inside the call `aes` inside the call to `ggplot`:

```{r tumour-lines-grouped-colour, exercise = TRUE, exercise.lines = 5}
tumours %>% 
  ggplot(aes(x = Time, y = TumourSize)) + 
  geom_point() + 
  geom_line(aes(group = TNO))
```

Now adapt that example by relocating the `col = TumourSize` part from the general `aes` call to the `aes` call inside `geom_line`.

```{r tumour-lines-grouped-colour2, exercise = TRUE, exercise.lines = 5}
tumours %>% 
  ggplot(aes(x = Time, y = TumourSize)) + 
  geom_point() + 
  geom_line(aes(group = TNO))
```

What happened?
Is that what you expected?
Do you understand how aesthetics can be inherited and overriden?


### Spider plot
We will finish off this session by producing a plot that is commonly seen in cancer trials literature.
A _spider plot_ shows the repeated measures of tumour size _relative to baseline_.
This last bit in italics is important because tumour sizes are analysed under the RECIST criteria for response compared to baseline.

To marry up each observation with the relevant baseline observation, we can use a `dplyr` join.

To do this, first note that the subset of rows with `Time == 0` is the set of baseline observations:

```{r}
tumours %>% 
  filter(Time == 0) %>% 
  select(TNO, Baseline = TumourSize)
```

Notice how the `TumourSize` variable in this dataset has been given the alias `Baseline`.
This is to avoid confusion in the imminent change-from-baseline calculation.

The baseline dataset can be joined to the original dataset by TNO:

```{r}
tumours %>% 
  left_join(
    tumours %>% 
      filter(Time == 0) %>% 
      select(TNO, Baseline = TumourSize), by = 'TNO'
  )
```

The join here:

* takes every row in the left-hand-side (LHS, here `tumours`)...
* and finds any number of matching records in the right-hand-side (RHS)...
* with the same `TNO` value...
* and pools the columns.

Note:
The significance of the word _left_ in `left_join` is that the LHS record is retained even if no match in the RHS could be found.
Other types of joins exist, like `right_join`, `full_join`, `anti-join`.
Be aware that the type of join determines what data you see. 
Joins can be confusing and deserve some dedicated study.
We close this sidebar on that cautionary note.

The joined data-set has TumourSize and Baseline in each record.
Obviously the two columns are equal when `Time` = 0.
The same is not true of later values - take a look by clicking on the blue tab numbers.

Before we can produce the spider-plot, we need to calculate the percentage change from baseline.
Do that now using the `mutate` command to calculate the `ChangeFromBaseLine` variable:

```{r tumours-change, exercise = TRUE, exercise.lines = 8}
tumours %>% 
  left_join(
    tumours %>% 
      filter(Time == 0) %>% 
      select(TNO, Baseline = TumourSize), by = 'TNO'
  ) %>% 
  mutate(ChangeFromBaseLine = )
```

Do your numbers look right?
If so, paste your command below and add commands from earlier examples in this section to produce line-plots of repeated measures by patient. 
Put `Time` on the $x$-axis, and `ChangeFromBaseLine` on the $y$-axis.

```{r tumours-change-plot, exercise = TRUE, exercise.lines = 11}

```

Before clicking through to see my attempt on the next page, attempt these embellishments:

* add some colour;
* add a horizontal line with `geom_hline` at $y$ = -30%. This is the RECIST threshold for _partial response_;
* facet to show the arms in separate plots
* add smoothers for each Arm;
* fade the patient-level lines using alpha = 0.25 (note this is not data-dependent so it does not go in `aes()`!)



## Final spider-plot

```{r, warning=FALSE, message=FALSE}
tumours %>% 
  left_join(
    tumours %>% 
      filter(Time == 0) %>% 
      select(TNO, Baseline = TumourSize), by = 'TNO'
  ) %>% 
  mutate(
    ChangeFromBaseline = 100 * (TumourSize / Baseline - 1),
    Arm = factor(Arm)
  ) %>% 
  ggplot(aes(x = Time, y = ChangeFromBaseline, col = Arm)) + 
  geom_point(size = 0.2) + 
  geom_line(aes(group = TNO), alpha = 0.25) + 
  geom_hline(yintercept = -30, col = 'red', linetype = 'dashed') + 
  geom_smooth() + 
  facet_wrap(~ Arm) + 
  labs(y = '% change from baseline')
```

How does your attempt compare to mine?

More importantly, which treatment would you prefer?


## Quiz
Let us recap some of what we have learned with a quiz.

```{r quiz, echo = FALSE}
quiz(
  
  question("What name identifies the abstract concept of associating aspects of a plot with your data values?",
    answer("Geometries", correct = FALSE),
    answer("Aesthetics", correct = TRUE),
    answer("Properties", correct = FALSE)
  ),
  
  question("What name identifies layers of a plot",
    answer("Geometries", correct = TRUE),
    answer("Aesthetics", correct = FALSE),
    answer("Properties", correct = FALSE)
  ),
  
  question("What name identifies the flexible parametric smoother we used?",
    answer("Lois", correct = FALSE),
    answer("LowEss", correct = FALSE),
    answer("Loess", correct = TRUE),
    answer("Louis", correct = FALSE),
    answer("Louise", correct = FALSE),
    answer("Dave", correct = FALSE)
  ),
  
  question("Can aesthetics be inherited?",
    answer("Yes", correct = TRUE),
    answer("No", correct = FALSE)
  ),
  
  question("What sequence or symbol do you use to add layers to a ggplot?",
    answer("&", correct = FALSE),
    answer("+", correct = TRUE),
    answer("*", correct = FALSE),
    answer("and also", correct = FALSE)
  ),
  
  question("Do you like ggplot?",
    answer("Yes", correct = TRUE),
    answer("No", correct = FALSE)
  )
)
```

## Exercise
Take the spider-plot example and adapt it to create a _waterfall plot_.
This type of plot, seen commonly in contemporary oncology trials research, shows the best change from baseline in tumour size in individual patients, ordered as a cascade from the patient showing worst change to the patient showing the best change. 
An example is shown below.

Some hints:

* The `dplyr` command `arrange(col_name)` can be used to order a data set by `col_name`;
* you can use `aes(x = seq_along(y_var), y = y_var)` to generate a sequence 1, 2, ..., `length(y_var)` for `x`;
* `geom_col` will produce a bar-chart from `x` and `y` aesthetics;
*  the `labs` command will allow you to override labels for the $x$- and $y$-axes, amongst others;
* the aesthetic `col` for a `geom_col` controls the colour of the line enclosing the bar; the aesthetic `fill` controls the colour used to fill the bar.

```{r, echo = FALSE}
tumours %>% 
  left_join(
    tumours %>% 
      filter(Time == 0) %>% 
      select(TNO, Baseline = TumourSize), by = 'TNO'
  ) %>% 
  mutate(
    ChangeFromBaseline = 100 * (TumourSize / Baseline - 1),
    Arm = factor(Arm)
  ) %>% 
  filter(Time > 0) %>%
  group_by(TNO, Arm) %>%
  summarise(BestChangeFromBL = min(ChangeFromBaseline)) %>% 
  arrange(-BestChangeFromBL) %>%
  ggplot(aes(x = seq_along(BestChangeFromBL), y = BestChangeFromBL, fill = Arm)) +
  geom_col() + 
  labs(y = 'Best % change from baseline', x = '')
```

See how close you can get to this image.
If you cannot reproduce it fully, do as much as you can, and make a note of the aspects you could not achieve.
Or better yet, try to improve on it.

Please email your image and code to k.brock@bham.ac.uk

