---
title: "Tutorial title"
author: "Joe Bloggs"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE)

# Any setup R code can go here
library(dplyr)
# Etc

```


## Welcome

This tutorial DOES WHAT?

### Learning Objectives
Upon completion of this session, you will:

* Be able to fulfill Learning Objective 1;
* Be able to fulfill Learning Objective 2;
* Be able to fulfill Learning Objective 3.


## Section 1

Introduce this section ...

Demonstrate some code:
```{r}
mtcars %>% head()
```

Set an exercise:
```{r exercise1, exercise = TRUE}

```

Set an exercise where you start the example:
```{r exercise2, exercise = TRUE}
mtcars %>% 
  filter(mpg < 20)

```

Set an exercise where you provide a hint:
```{r exercise3, exercise = TRUE}

```
```{r exercise3-hint}
mtcars %>% 
  filter(mpg < 20) %>% 
  group_by(cyl) %>% 
  summarise(mean(hp))
```



## Quiz
Throw in some quizes to check learning progress:

```{r quiz, echo = FALSE}
quiz(
  
  question("Which of the following are animals?",
    answer("Cat", correct = TRUE),
    answer("Petunia", correct = FALSE),
    answer("Turmeric", correct = FALSE),
    answer("Lizard", correct = TRUE)
  ),
  
  question("Reviewers that demand p-values should be:",
    answer("Rewarded with cuddles.", correct = FALSE),
    answer("Fired into the sun.", correct = TRUE)
  )
)
```


