---
title: "More tidyverse"
author: "Kristian Brock"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE)

# Any setup R code can go here
# Etc
```


## Welcome

This tutorial introduces further elements of the `tidyverse` suite of packages, specifically `dplyr`, `tidyr` and `lubridate`.

### Learning Objectives
Upon completion of this session, you will:

* Be able to use different types of join in `dplyr`;
* Be able to reshape data using `tidyr`;
* Be able to calculate the time between two dates using `lubridate`.


## Joins in `dplyr`

Joins combine two datasets together by matching one or more columns in both the left- and right-hand objects, returning the pooled set of columns.
Thus, one way to interpret a join is an operation that can make a dataset wider by adding more columns.
For instance, we already used joins in an earlier bootcamp session to create spider plots.

Let us remind ourselves of the `tumours` dataset:
```{r}
library(bootcamp)
library(dplyr)

tumours %>% head()
```

In the spider plot exercise, we joined `tumours` to a subset of itself to add the baseline tumour size to all records.
This allowed us to calculate change-from-baseline at all timepoints.

In addition to `tumours`, the `bootcamp` package also contains a `patients` dataset containing qualitative information on patients in a putative trial:

```{r}
patients %>% head()
```

Notice that both datasets contain a `TNO` record?
To join `tumours` to `patients`, we can run:

```{r}
tumours %>% 
  inner_join(patients, by = 'TNO')
```

Here we used `inner_join` to join the left-hand side (LHS) `tumours` to the right-hand side (RHS) `patients`, via the _join key_ `TNO`.
There are actually several different types of join:

* an `inner_join` returns the set of records where the key is present in both the LHS and RHS;
* a `left_join` returns all records in LHS, adding values from the RHS where a key match is available or a row of `NA` values where no match is available;
* a `right_join` returns all records in RHS, adding values from the LHS where a key match is available or a row of `NA` values where no match is available;
* a `full_join` returns all records in the LHS and RHS, adding values from the other side where a key match is available or a row of `NA` values where no match is available;
* a `semi_join` returns all records in the LHS where a key match is available in the RHS, retaining only the columns in the LHS;
* an `anti_join` returns all records in the LHS where no key match is available in the RHS, retaining only the columns in the LHS;

Notice that:

* An inner-join is like the intersection of two sets;
* An outer-join is like the union of two sets;
* left and right joins are a way of nominating a master set of records;
* semi- and anti-joins can be interpreted as ways of filtering; 
* semi- and anti-joins bring no extra variables;
* semi- and anti-joins are opposites

All of these joins use the pattern:

```{r, eval=FALSE}
lhs %>% 
  join(rhs, by = 'Key')
```

Write a join to return patients with no tumour measurement:
```{r exercise1, exercise = TRUE}

```

Sometimes the join key may have different names in the LHS and RHS.
When this happens, use:

```{r, eval=FALSE}
lhs %>% 
  join(rhs, by = c('ColInLHS' = 'ColInRHS'))
```

Sometimes the join key may include several columns.
When this happens, use:

```{r, eval=FALSE}
lhs %>% 
  join(rhs, by = c('Col1InLHS' = 'Col1InRHS',
                   'Col2InLHS' = 'Col2InRHS'))
```



## Reshaping data using `tidyr`

The `tidyr` package provides many useful functions with reshaping and tidying data.
The two functions we will learn about in this session convert between wide and tall data formats.

### Tall to wide

The `tumours` dataset is said to be _tall_ because each row pertains to exactly one measurement event.
This data format is require by `ggplot` to produce plots, and functions in R to fit linear models like $$TumourSize = \alpha + \beta Time + \epsilon$$.

Although computers prefer the tall format, humans tend to prefer the _wide_ format, where 

This can be created using the `spread` function in the `tidyr` package:

```{r}
library(tidyr)

tumours %>% 
  spread(Time, TumourSize)
```

This commands says:

* create a column for each value of `Time`
* containing the values in `TumourSize`
* whilst treating all other columns as fixed.

If you were asked to tabulate the raw values by patient, this would likely be much more useful to a reviewer than the tall format.
For instance, we can immediately observe by looking across the rows that most patients have tumour shrinkage.

Thus, `spread` takes you from tall to wide data.

As an exercise, reproduce the following tabulation of the mean tumour size by `Arm` and `Time` using the `group_by`, `summarise` and `spread` commands:

```{r, echo = FALSE}
tumours %>% 
  group_by(Arm, Time) %>% 
  summarise(MeanSize = mean(TumourSize)) %>% 
  spread(Time, MeanSize)
```

```{r exercise3, exercise = TRUE}



```

### Wide to tall
The opposite problem concerns going from a wide to a tall format.
This frequently occurs when researchers send you data to analyse.
They will generally have recorded data in a spreadsheet in the wide format, like this:

```{r}
tumours %>% 
  spread(Time, TumourSize) -> tumours_wide

tumours_wide
```

To visualise or analyse the data, you will likely have to convert it from wide to tall format first.
To do this, use the `gather` funciton in `tidyr`:

```{r}
tumours_wide %>% 
  gather(Time, TumourSize, -TNO, -Arm)
```

This command says:

* take the column headers from `tumours_wide` and create a new column called `Time`
* insert the table values into a new column called `TumourSize`
* treat `TNO` and `Arm` as identity fields that should not be stacked.



## Working with dates in `lubridate`

`lubridate` is a package in `tidyverse` that aims to take the pain out of working with dates in R.

```{r}
library(lubridate)
```

You can calculate the duration of time between two dates by using the `interval` function:
```{r}
patients %>% 
  mutate(AgeAtReg = interval(DateOfBirth, DateOfReg))
```

Notice how the `AgeAtReg` variable is not yet telling us anything useful?
It is in the default format of an interval.
What we actually want is the number of years in that interval.
To do that, we divide the interval by a single year:

```{r}
patients %>% 
  mutate(AgeAtReg = interval(DateOfBirth, DateOfReg) / years(1))
```

Ages are one of the few variables that are generally always rounded downwards.
Hence, if you had asked each of these patients at their regostration "How old are you?", they would have replied:

```{r}
patients %>% 
  mutate(AgeAtReg = floor(interval(DateOfBirth, DateOfReg) / years(1)))
```

You can create a date object from a string using the `ymd` parser function:

```{r}
halloween <- ymd('2019-10-31')
class(halloween)
halloween
```

Mutate the `patients` dataframe to show the number of days between Halloween and the date each patient registered:
```{r exercise4, exercise = TRUE}



```

Calculate the median interval by `StarSign`, in days:
```{r exercise5, exercise = TRUE}





```

What is the median follow-up for Geminis?


There are lots more useful functions offered by the `dplyr`, `tidyr` and `lubridate` packages.
This session introduced methods to perform three very common tasks:

* joining datasets
* transitioning between wide and tall data formats
* and calculating length of time intervals

There are also lots more useful packages in the `tidyverse`.
Keep exploring.
