---
title: "Linear models in R"
author: "Kristian Brock"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE)
```

## Linear models

Ordinary least squares regression models can be run using the R command `lm` for _linear model_.
Generalised linear models (e.g. logit) are provided by the `glm` command.
Multi-level linear models (e.g. patient-level intercepts) are provided by the `nlme` and `lme4` packages, and `lme4` also suppports multi-level generalised linear models.

```{r}
model_data <- data.frame(
  TNO = c(1, 1, 1, 2, 2, 2, 3, 3, 4, 4, 4),
  Arm = c(0, 0, 0, 1, 1, 1, 0, 0, 1, 1, 1) %>% as.factor,
  Time = c(0, 3, 6, 0, 3, 6, 0, 6, 0, 3, 6),
  TS = c(10, 8, 7, 8, 6, 4, 7, 8, 4, 3, 1)
)

model_data
```

The data looks like this:
```{r, echo = FALSE}
library(ggplot2)

model_data %>% 
  ggplot(aes(x = Time, y = TS, group = TNO, col = Arm)) + 
  geom_point() + 
  geom_line() + 
  theme(legend.position = 'bottom')
```

We have four patients with up to 3 measures of `TS` (tumour size) in each. 
Let us assume that we want to analyse changes in `TS`.
We could fit a linear model.
```{r}
lm1 <- lm(TS ~ 1 + Time, data = model_data)
```

The `~` sign says that the thing on the left is estimated by the things on the right.
On the right of `~` appear the explanatory variables.
The `1` stands for an intercept, or the average response value when all other explanatory variables are zero.
`Time` is also included as a covariate.
By default,  `Time` is treated as a continuous quantity.

We have run the model and assigned the result to the variable named `lm1`.
There are many functions we can call on `lm1` to access useful information.
For instance, to get a quick summary of the model, we can run:

```{r}
summary(lm1)
```

We see that the average intercept is about 7 and that `TS` decreases on average by about 0.4 units per unit time.

We can fetch the fitted values with `fitted`.
Let's plot them against `Time`.
```{r}
x <- model_data$Time
y <- fitted(lm1)
plot(x, y, type = 'b')
```
